### 2024-08-12 Paper/paper2.tex
Updates from Overleaf

	\linespread{1.2}\title{\vspace{-0.5in} Does Hospital Leadership Matter?\\ \large Evidence from Pay-for-Performance Incentives}

@@

    \begin{abstract}
		{\small
         One determinant of firm behaviors and performance is the type of management leading the firm. Characteristics such as age, gender, and background are correlated with employee wages, decisions of mergers and acquisitions, and other firm outcomes. In this paper, I investigate this topic in the context of hospital executive teams. Specifically, I estimate the effect of clinically trained executives on hospital quality. I leverage pay-for-performance policies that changed quality incentives for hospitals in the US in 2012, and estimate the difference in readmission and mortality rates for nonprofit hospitals with and without clinically trained executives. I find that clinically trained executives do not improve quality as much as executive teams without clinically trained executives. This difference is not driven by a change in patient complexity, but may be driven by preferences towards profit vs. patient welfare. Further, clinically trained executives are not just a signal of underlying hospital incentives, but manage the hospital differently. 
		} 
	\end{abstract}

	\vspace{0.8in}

	\noindent Keywords:

	\noindent JEL Codes:

@@

    Broadly, firm leader characteristics such as gender, age, and background have been shown to be correlated with financial performance and behaviors such as employee diversity, wages, aggressiveness in mergers and acquisitions, and other corporate strategies. I consider this topic in the context of hospital quality under executives with differing backgrounds: some with training as a medical doctor and some without. Physicians can bring a unique combination of clinical and administration expertise, and they have insights that can potentially benefit their employees and lead to better quality (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). However, hiring physician executives can also be controversial, as physicians are not always adequately trained in business management, which can lead to poor business practices and lower quality (\cite{HarvardBusinessReview2018}).

    In this paper, I estimate how having clinically trained executives in leadership affects quality of care in hospitals. I scrape executive team names and titles from publicly available Tax Form 990s, filled out by nonprofit firms in the US. Combining this with publicly available hospital characteristics, I construct a novel data set of nonprofit hospitals from 2010-2014 containing information on both executive team characteristics and quality metrics. While limited to nonprofit hospitals due to data availability, this sample is a very important driver of health care in the US. Private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up 36\% of hospitals and staff 107 beds (\cite{ASPE_2023}). Therefore, the behavior of these hospitals directly affects health care consumers in the US.

    The non-random selection of leaders makes it difficult to interpret any direct comparison of differing executive teams as causal. Therefore, I leverage an exogenous shock to hospital incentives, pay-for-performance policies enacted in the US in 2012, and compare the change in readmission and mortality rates of executive teams with and without clinical executives. Under the assumption that the two types of teams would have had similar trends in readmission and mortality rates absent the incentive change, I find that executive teams with clinical executives on average do not improve quality as much as executive teams without clinical executives. While both types of hospital leadership teams decrease readmission rates after the policy change, those without clinical executives decrease readmissions at a faster rate than those with clinical executives.

    Having found that hospital leadership does, in fact, matter to hospital behavior, I investigate several mechanisms that might contribute to this difference. First, the more drastic decrease in readmissions by non-clinical teams is not driven by decreasing average patient complexity. Further, comparing each type of leadership team with for-profit hospitals reveals that the absence of clinical executives leads a nonprofit hospital to respond more similarly to a for-profit hospital than nonprofits with clinical executives. That is, having clinical training on a leadership team leads a hospital to act less profit-driven than their general training counterparts. Finally, I show that changes in a hospital's propensity to hire a clinical executive are not correlated with the pay-for-performance policies. Thus, I exploit such changes in leadership teams to decompose the effect of clinical executives into signaling (clinical executives signal underlying hospital preferences), or managing (clinical executives manage the hospital differently). I find that the entire difference in response is driven by a managing effect.

@@

@@

    Nonprofit firms in the US differ from for-profits mainly in that for-profit firms distribute revenue to stakeholders, while nonprofits reinvest profit back into the firm. Hospitals are governed by a board of directors, whose role is to set broad goals and strategies, and provide general oversight. The board selects executives, the highest level of management of a firm, to carry out day-to-day operations. An executive team usually consists of at least a Chief Executive Officer (CEO) and Chief Financial Officer (CFO), but there is variation in how firms organize executive teams. Some hospital executives specialize in health care settings by earning a degree in health care management, or an MBA specific to health care. There are executive positions that are often filled by someone with clinical training, such as a Chief Medical Officer (CMO), but medical doctors can pursue other top managerial positions as well such as CEO or president. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive.

    Hospital executive teams are understudied partly due to the inaccessibility of granular data. I construct a novel data set of nonprofit hospitals in the US which contains names, titles, and positions of all board members and executives tied to the hospital from 2009-2015. I gather this data from publicly available Tax Form 990s, which every sufficiently large nonprofit files each year in the US. To my knowledge, this is the first large scale gathering of executive names from these forms.\footnote{\citeauthor{brickley2010board} (\citeyear{brickley2010board}) collects compensation data for a select number of hospitals.}

    All tax-exempt organizations in the US are required to file a Form 990 with Internal Revenue Services (IRS) each year. There are different types of forms, but any organization grossing over \$200,000 must file the most extensive Form 990. Sections of this form include a statement of revenue, statement of functional expenses, a balance sheet, and, as used in this project, a list of all key employees, executives, and board members. Each firm is required to report the name and title, average hours per week, position, and compensation of their board members and executives.

    The historical tax forms are housed by ProPublica.\footnote{https://projects.propublica.org/not-for-profits/} I use the NonProfit Explorer API to extract Employee Identification Numbers (EIN) of firms with a hospital National Taxonomy of Exempt Entities (NTEE) code. After filtering out associations and specialty hospitals, there are approximately 3,000 EINs. Linking these EINs to other hospital information is crucial to the analysis, but no crosswalk from EIN to other hospital identifiers is publicly available. I therefore rely on matching by location and name of hospital in both the tax forms and American Hospital Association (AHA) data. In the AHA data, I limit to general hospitals. Then, I confidently match 1200 EINs to an AHA identification number based on exact name matches within the same state.\footnote{I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.}

    For each of the matched EINs, I extract web URLs to Form 990 PDFs for each year. I use an algorithm to download these locally and use optical character recognition (OCR) text extraction methods to scrape the relevant section of the PDF, which is titled ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees". Finally, I use string cleaning methods to identify all names and positions under each EIN, year. With each name, I extract titles indicating clinical training: MD, Dr., or DO. Using OCR extraction can be unreliable in cases where the text on the form is handwritten or small. Therefore, for observations that are missing after the initial extraction, I manually record names, positions, and titles. As I limit the sample to hospitals who have text information in every year, I limit the years to 2010-2014 to retain more hospitals. This yields 852 nonprofit hospitals that are matched to AHA data, and with complete text information in each year.

    Finally, I seek to (1) verify clinical training of those who claim to be doctors, and (2) obtain additional information about clinically trained executives. Thus, I match each name with the National Plan and Provider Enumeration System (NPPES) database of all registered physicians. For any person who has a doctor title and uniquely matches the NPPES data, I record their National Provider Identifier (NPI) number. For anyone who has a doctor title and no matches, no doctor title and matches, or a doctor title and multiple matches, I use information on location and hospital name to manually match NPI numbers of those who did have clinical training. I present a table of means for various characteristics of doctor executives in the data in Table \ref{doc_sumstats}. In total, there are 813 doctor executives. On average, they are 52 years old, and 11\% of them are female. While it is more likely they are a Chief Medical Officer than a Chief Executive Officer, still less than a third are CMOs. The most common specialty is internal medicine.

@@

    For the purposes of this paper, I focus on executive teams as a whole instead of specific positions. The main reason for this is because the organization of teams varies so drastically across hospitals. It is not uncommon for two hospitals to have the same position but the tasks performed are drastically different, or to have two distinct positions that essentially perform the same tasks. I drop anyone who is solely a board member, and focus on those who have executive in their title, or are labeled as a president or vice president of a department. I create various hospital-level characterizations based on their executive team, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer. For the main analysis, I limit to hospitals who either always have a clinical executive or never have a clinical executive over the course of the sample in order to have a clean comparison.

@@

    Two programs were passed as part of the Affordable Care Act that focused on pay-for-performance incentives for hospitals: the Hospital Readmissions Reduction Program (HRRP), and the Hospital Value Based Purchasing Program (HVBP). HRRP focused on penalizing hospitals with poor quality, and HVBP focused on rewarding hospitals with high quality or show improvement in quality.

@@

    I focus on outcomes of readmission and mortality rates, which are publicly available at the hospital-condition level in CMS Hospital Compare. I combine rates for pneumonia, AMI, and heart failure (the relevant HRRP conditions) into a weighted average based on the number of patients in each condition. Additionally, I include several relevant hospital characteristics from the AHA survey: number of beds, and whether the hospital is an academic medical center, is physician owned, or is system affiliated. Additionally, case mix index, a measure of overall patient complexity, comes from the CMS Case Mix Index files. Whether the hospital was penalized under HRRP or received payments under HVBP is found in the HCRIS data.

@@

    The hospitals in my sample are limited to general nonprofit hospitals with executive team information in each year (2010-2014) and a link from EIN to AHA ID. In Table \ref{tab:sumstats_samples_stable}, I present means of relevant variables for several hospital samples used throughout the paper. First, in column (1), are means for hospitals who ever have a clinical executive at any point in the sample. In column (2) are means for hospitals who have a clinical executive for the entire sample period, and column (3) shows means for hospitals who never have a clinical executive in the sample period. The main analysis compares hospitals in columns (2) and (3), and supplemental analyses leverage hospitals in column (1).

@@

    Hospitals that never have a clinical executive are smaller, less likely to be an academic medical center, and less likely to be affiliated with a system. Further, they have smaller executive teams on average by 1-2 people, and are less likely to have anyone acting as a CMO.\footnote{Though, they may title the position differently.} Finally, in regards to exposure to pay-for-performance programs, hospitals who never have clinical executives are less likely to receive benefits from the HVBP Program, but seem more similar to hospitals with clinical leadership in terms of penalties from HRRP.

    Next, I plot weighted average readmission, mortality rates, and patient case mix index across time for different hospital types in Figure \ref{fig:outcomes_graph}. Pay-for-performance incentive change timing is shown as a dotted line in 2012. Average readmission rates trend similarly prior to 2012, and all hospital type exhibit a decrease in readmission after 2012. However, hospitals who consistently have a clinical executive show a slower rate of decrease in raw readmissions. When considering mortality, there is not a drastic change in 2012 for any hospital type, but hospitals who always have a clinical executive continue on an increasing trend in mortality where the other hospitals types remain more flat. Finally, hospitals that never have clinical executives consistently have less complex patients, but there is no notable change in raw case mix index over time.

@@

    \section{Effect of Executive Clinical Training on Hospital Quality}\label{sec:clinical}

    First, I test empirically whether having clinical training matters in how hospitals change readmission and mortality rates in response to the pay-for-performance policies effective in 2012. For a clean comparison, I limit the sample to hospitals who always have clinical executives or never have clinical executives. That is, these hospitals do not change their propensity to hire a clinical executive during the sample period. Therefore, an important assumption that I make is that this sample selection is not correlated with the program enactment. I investigate this assumption in Section \ref{sec:changes}, and find no evidence that hospitals are more or less likely to change their propensity to hire a clinical executive because of the pay-for-performance programs.

@@

    In the unweighted two-way-fixed-effects specification, we have reason to believe that the parallel trends assumption may not hold due to observed differences in characteristics other than the leadership team, as discussed in Section \ref{sec:data}. Once incorporating the weights used in the synthetic differences in differences method, we ensure parallel trends in the outcome variables prior to 2012. Thus, I assume that under the weighted clinical and non-clinical hospital composition, trends would have continued to be parallel absent the incentive change in 2012. A component of this assumption is the implicit assumption that no other event occurred in 2012 correlated with both the outcomes and composition of leadership team. Finally, I assume that there is no anticipation of the policy change before 2012. The rules of the policies were announced in October of 2011, so 2012 was the first full year that hospitals could meaningfully respond to the policies.

    The estimated difference in readmission and mortality rates between hospitals with and without clinical executives are shown in Figure \ref{fig:clinicalsynthdid}. Estimate $\beta$ is represented in the difference in slope after 2012 when pay-for-performance incentives began. Panel \ref{fig:read_synth_clinical} shows the estimated difference in readmission rates, and reveals that non-clinical teams decrease readmissions by .3 (standard error .13) more than clinical teams. This is a relatively small magnitude given an average readmission rate of 21, but reveals a meaningful difference in behavior given that both hospital types significantly decrease readmissions. In panel \ref{fig:mort_synth_clinical}, I show the estimated difference in mortality. While the magnitude of the difference is similar to that for readmission rates, mortality is a noisy measure, and I cannot rule out no difference in mortality response. In Appendix \ref{app:condition}, I show that these results are primarily driven by pneumonia patients.

     \begin{figure}[ht!]
     \caption{Effect of Clinical Experience on Readmission and Mortality Rates}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{Readmission Rate}
         \includegraphics[width=\textwidth]{Objects/read_md_nomd_synth_graph.pdf}
         \label{fig:read_synth_clinical}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{Mortality Rate}
         \includegraphics[width=\textwidth]{Objects/mort_md_nomd_synth_graph.pdf}
         \label{fig:mort_synth_clinical}
     \end{subfigure}
        \label{fig:clinicalsynthdid}
    \end{figure}

    Additionally, I estimate the intensive margin of clinical leadership: whether the fraction of clinical executives on the team affects quality. I bin the fraction of clinical executives into less than a quarter or more than a quarter and less than one half.\footnote{Less than .5\% of hospitals have greater than 50\% clinical executives on the team.} Then, I estimate:

    \begin{equation}
    \label{eq:clinical_continuous_onefourth}
    y_{ht} = \beta \left(\mathbbm{1}\{0 < \rho \leq .25\}\text{ x post 2012}\right)_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}, \text{ and}
    \end{equation}
    \begin{equation}
    \label{eq:clinical_continuous_onehalf}
    y_{ht} = \beta \left(\mathbbm{1}\{.25 < \rho \leq .5\}\text{ x post 2012}\right)_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht},
    \end{equation}

@@

    I show the results of this estimation with readmission rates as the outcome in Figure \ref{fig:cont_clinicalsynthdid}. In panel \ref{fig:onefourth_read_synth_clinical}, the estimation compares executive teams with no clinical experience to those with up to 25\% of the team having clinical experience. The average treatment effect is .04 and not statistically significant. Thus, there is no significant difference in how these types of teams respond to the incentive change. In panel \ref{fig:onehalf_read_synth_clinical}, I compare hospitals with no clinically trained executives over the course of the sample to hospitals with over a quarter of their team having clinical training. While these hospitals have parallel pre-trends, hospitals without clinically trained executives decrease readmissions by .64 (standard error .33) more than hospitals with clinically trained executives. Thus, the readmission rate response is fully driven by having a higher proportion of clinical executives relative to no clinical executives.

    \begin{figure}[ht!]
     \caption{Effect of Clinical Training on Readmission Rates, Binned Treatment}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{$0 < \rho \leq .25$}
         \includegraphics[width=\textwidth]{Objects/cont_onefourthread_md_nomd_synth_graph.pdf}
         \label{fig:onefourth_read_synth_clinical}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{$.25 < \rho \leq .5$}
         \includegraphics[width=\textwidth]{Objects/cont_onehalfread_md_nomd_synth_graph.pdf}
         \label{fig:onehalf_read_synth_clinical}
     \end{subfigure}
        \label{fig:cont_clinicalsynthdid}
    \end{figure}

    Any potential difference in mortality rate response also seems to be driven by executive teams with over a quarter of the executives having clinical experience, as shown in Figure \ref{fig:cont_clinicalsynthdid_mort}. However, the standard errors of this estimation are high, potentially due to the noisiness in the mortality measure. The magnitude of the estimate in panel $\ref{fig:onefourth_mort_synth_clinical}$ is .07, so I can rule out differential responses by teams with less than a quarter of clinical executives. Taking the readmission and mortality rate results together, the effect of having clinical leadership on quality after an incentive change depends on having a sufficient proportion of the team with clinical training.

>>>>

	\linespread{1.2}\title{\vspace{-0.5in} Does Hospital Leadership Matter?\\ \large Evidence from Pay-for-Performance}

@@

    \begin{abstract}
		{\small
         Managers are often seen as drivers of firm decisions, strategies, and performance, and manager heterogeneity has been linked to employee wages, decisions of mergers and acquisitions, and financial performance. A significant goal in US health care is to improve quality while decreasing spending, and yet it is unclear whether hospital managers might contribute to this goal. In this paper, I estimate the effect of clinically trained executives on hospital quality. I leverage pay-for-performance initiatives that changed financial incentives on quality of care for hospitals in the US in 2012, and estimate the difference in readmission and mortality rates for nonprofit hospitals with and without clinically trained executives. I find that hospitals without clinically trained executives respond more to financial incentives than those without clinically trained executives. This difference is not driven by a change in patient complexity, but may be driven by preferences towards profit vs. patient welfare.  
		} 
	\end{abstract}

@@

    \section{Introduction}

    Executives are considered crucial drivers of a company's practices and performance, as they often set broad visions and implement strategies to achieve them. Executive heterogeneity has been linked to various decisions and outcomes of firms: characteristics such as gender, age, and background are correlated with financial performance and behaviors such as employee diversity, wages, aggressiveness in mergers and acquisitions, and other corporate strategies.\footnote{\cite{bertrand2003managing}; \cite{matsa2011chipping}; \cite{matsa2011chipping}; \cite{adams2009women}; \cite{ahern2012changing}; \cite{flabbi2019female}; \cite{benmelech2015military}; \cite{custodio2013generalists}; \cite{frydman2019rising}} However, there is still much to learn about how firm behaviors and objectives can be shaped by differences in leadership.

    In this paper, I study how executive background affects firm behavior in the context of US hospitals. In the US, health care makes up almost 20\% of spending (\cite{CMS_National_Health_Expenditure}), and a major goal of policymakers is to improve the value of health care. Yet, little is known about how hospital managers affect health care value, and it is unclear whether existing findings on executive characteristics can be extrapolated to the health care context. This is because existing literature is generally limited to large, publicly traded firms, whereas a majority of US hospitals are nonprofit. Relatedly, health care executives have a particularly unique trade-off between profit and patient welfare, which varies from the types of firms studied in existing literature.

    A characteristic relevant to this context is whether or not hospital executives have clinical training. Physicians can bring a unique combination of clinical and administration expertise, and they have insights that can potentially benefit their employees and lead to better quality (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). However, hiring physician executives is controversial, as physicians are not always trained in business management, which can lead to poor practices and lower quality (\cite{HarvardBusinessReview2018}). Thus, I estimate how having clinically trained executives in hospital leadership affects quality of care. I scrape executive names and titles from publicly available Tax Form 990s, filled out by nonprofit firms in the US. Combining this with publicly available hospital characteristics, I construct a novel data set of nonprofit hospitals from 2010-2014 containing information on both executive team characteristics and quality metrics. While limited to nonprofits due to data availability, this sample of hospitals is a very important driver of health care in the US. Private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up 36\% of hospitals and staff 107 beds on average (\cite{ASPE_2023}).

    The non-random selection of leaders makes it difficult to interpret any direct comparison of differing executive teams as causal. Therefore, I leverage pay-for-performance policies enacted in the US as an exogenous shock to hospital incentives, and compare the change in readmission and mortality rates of hospitals with and without clinically trained executives over the entire sample period. Under the assumption that the two types of teams would have had similar trends in readmission and mortality rates absent the incentive change, I find that hospitals without clinically trained trained executives respond more to financial incentives on quality than hospitals with clinical executives. That is, while both types of hospital leadership teams decrease readmission rates after the policy change, those without clinical executives decrease readmissions at a faster rate than those with clinical executives. Interestingly, this result is driven by hospitals with between one quarter and one half of their executive team having clinical training. Further, executives with clinical training in fields specific to the policy change respond more than other clinically trained executives.

    One potential reason for this differential response is that hospitals without clinically trained executives are more willing to participate in selective patient practices. To investigate this, I use patient case mix index as an outcome and estimate the difference in case mix index after the financial incentive change. I find no evidence that non-clinical executive teams differentially change patient complexity.

    Next, I investigate whether this differential response could be due to differences in objective functions. I consider hospital objectives as a function of profit and patient welfare, and allow weight placed on each to vary with executive team type. Hospitals that place more weight on profit will naturally respond more drastically to a policy that adds quality directly into the profit function. Thus, a potential reason for the difference in response is that non-clinical executive teams are more profit driven. I investigate this empirically by comparing both types of hospital leadership teams to for-profit hospitals, who have revealed that they are largely profit driven by their ownership type. This analysis reveals that non-clinical executive teams respond to the financial incentives similarly to for-profits, and hospitals with clinically trained executives act differently than for-profits.

    Finally, the question still remains whether clinically trained executives are a signal of underlying hospital preferences, or if they actively manage the hospital differently than their non-clinically trained counterparts. I analyze this by leveraging changes in hospital propensity to hire clinical executives to decompose the full effect of clinical leadership into a signaling effect (clinical executives signal underlying hospital preferences), or managing effect (clinical executives manage the hospital differently). First, I show that hospital executive changes are not correlated with the change to financial incentives. For the signaling effect, I compare hospitals who ever have a clinically trained executive to hospitals who never have a clinically trained executive. For the managing effect, I limit to hospitals who ever have a clinically trained executive and compare those who have a clinically trained executive when the policies were announced to those who have clinically trained executives at other times. I find that the difference in response is driven by a managing effect.

@@

    I use several data sets to analyze the relationship between hospital executives with clinical training and response to financial incentives on quality. I construct a novel data set of hospital executives by scraping publicly available tax forms, and merge this data to publicly available data sets containing hospital characteristics: the American Hospital Association Survey (AHA), the Centers for Medicare and Medicaid Services (CMS) Hospital Compare data and Case Mix Index files, and Healthcare Cost Report Information System (HCRIS) data. Ultimately, I employ this hospital-level data spanning from 2010-2014, which contains executive team characteristics, mortality and readmission rates as measures of quality, and other characteristics. A detailed description of the construction of this data can be found in Appendix \ref{appendixdata}.

@@

    Nonprofit hospitals, as with most nonprofits, are typically governed by a board of directors, whose role is to set broad goals and strategies, and provide general oversight. The board selects executives, the highest level of management of a firm, to carry out day-to-day operations. An executive team usually consists of at least a Chief Executive Officer (CEO) and Chief Financial Officer (CFO), but there is variation in how firms organize executive teams. Some hospital executives specialize in health care settings by earning a degree in health care management, or an MBA specific to health care. There are executive positions that are often filled by someone with clinical training, such as a Chief Medical Officer (CMO), but medical doctors can also fill other executive positions. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive.

    Hospital executive teams are understudied partly due to the inaccessibility of granular data. I construct a novel data set of nonprofit hospitals in the US which contains names, titles, and positions of all board members and executives tied to the hospital from 2009-2015. I gather this data from publicly available Tax Form 990s, which every sufficiently large nonprofit files each year in the US. To my knowledge, this is the first large scale gathering of executive names from these forms.\footnote{\citeauthor{brickley2010board} (\citeyear{brickley2010board}) collects compensation data for a select number of hospitals.} All tax-exempt organizations in the US are required to file a Form 990 with Internal Revenue Services (IRS) each year. There are different types of forms, but any organization grossing over \$200,000 must file the most extensive Form 990. Sections of this form include a statement of revenue, statement of functional expenses, a balance sheet, and, as used in this project, a list of all key employees, executives, and board members.

    For each hospital in the tax form data, I seek matches in the AHA data based on hospital name and location.\footnote{I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.} For all hospitals that can be matched to a general hospital in the AHA data, I use optimal character recognition (OCR) methods to extract the names, titles, and positions of each listed executive from its Tax Form 990. While board members are an important aspect of hospital organization, I do not include them in this analysis as they are beyond the scope of this paper. While I extract text spanning 2009-2015, I retain a much larger sample of hospitals by limiting to hospitals that have text information in 2010-2014. This yields 852 nonprofit hospitals that are matched to AHA data, and with complete text information in each year.

    Finally, I seek to (1) verify clinical training of those who are identified as doctors, and (2) obtain additional information about clinically trained executives. Thus, I match each name with the National Plan and Provider Enumeration System (NPPES) database of all registered physicians. I present a table of means for various characteristics of doctor executives in the data in Table \ref{doc_sumstats}. In total, there are 813 clinically trained executives. On average, they are 52 years old, and 11\% of clinically trained executives are female. While it is more likely they are a Chief Medical Officer than a Chief Executive Officer, still less than a third are CMOs. The most common specialty is internal medicine.

@@

    For the purposes of this paper, I focus on executive teams as a whole instead of specific positions. The main reason for this is because the organization of teams varies so drastically across hospitals. It is not uncommon for two hospitals to have the same position but the tasks performed are drastically different, or to have two distinct positions that essentially perform the same tasks. I create various hospital-level characterizations based on their executive team, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer.

@@

    The purpose of this paper is to investigate hospital behavior in the midst of changes to financial incentives that change how hospitals receive payment for quality of care. Two programs were passed as part of the Affordable Care Act that focused on pay-for-performance incentives for hospitals: the Hospital Readmissions Reduction Program (HRRP), and the Hospital Value Based Purchasing Program (HVBP). HRRP focused on penalizing hospitals with poor quality, and HVBP focused on rewarding hospitals with high quality or show improvement in quality.

    In October 2011, the Center for Medicare and Medicaid Services (CMS) released a set of rules under HRRP mandating penalties for hospitals with above average readmission rates. The goal of HRRP is to lower readmissions through better care coordination, less initial stay complications, and better post-care instructions. Beginning in October 2012, hospitals with higher readmission rates than the national average in pneumonia, heart failure, or AMI (after adjusting for demographic characteristics) receive a fixed lower reimbursement rate for all Medicare patients seen in their hospital. Penalties are given in the form of a fixed rate reduction of 1-3\% for every Medicare patient regardless of the condition. Excess readmission rates are calculated using a rolling look-back period of 3 years to determine whether the hospital is penalized. Therefore, hospitals had incentive to react immediately once details of the program were announced in October of 2011.

@@

    I focus on outcomes of readmission and mortality rates, which are publicly available at the hospital-condition level in CMS Hospital Compare. I combine rates for pneumonia, AMI, and heart failure (the relevant HRRP conditions) into a weighted average based on the number of patients in each condition.

@@

    In Table \ref{tab:sumstats_samples_stable}, I present means of relevant variables for several hospital samples used throughout the paper. First, in column (1), are means for hospitals who ever have a clinical executive at any point in the sample. In column (2) are means for hospitals who have a clinical executive for the entire sample period, and column (3) shows means for hospitals who never have a clinical executive in the sample period. The main analysis compares hospitals in columns (2) and (3), and supplemental analyses leverage hospitals in column (1).

@@

    Comparing columns (2) and (3), hospitals that never have a clinical executive are smaller, less likely to be an academic medical center, and less likely to be affiliated with a system. Further, they have smaller executive teams on average by 1-2 people, and are less likely to have anyone acting as a designated CMO. Finally, in regards to exposure to pay-for-performance programs, hospitals who never have clinical executives are less likely to receive benefits from the HVBP Program, but seem more similar to hospitals with clinical leadership in terms of penalties from HRRP.

    Next, I plot weighted average readmission, mortality rates, and patient case mix index across time for different hospital types in Figure \ref{fig:outcomes_graph}. The effective start year of pay-for-performance incentives is shown as a dotted line in 2012. Average readmission rates trend similarly prior to 2012, and all hospital types exhibit a decrease in raw readmission rates after 2012. However, there is a faster rate of decrease for hospitals who have a clinically trained executive at some point, or who never have a clinically trained executive. When considering mortality, there is not a drastic change in 2012 for any hospital type, but hospitals who always have a clinical executive continue on an increasing trend in mortality where the other hospitals types remain more flat. Finally, hospitals that never have clinical executives consistently have less complex patients, but there is no notable change in raw case mix index over time.

@@

    \section{Effect of Executive Clinical Training on Response to Quality Incentives}\label{sec:clinical}

    First, I test empirically whether having clinical training matters in how hospitals change readmission and mortality rates in response to the pay-for-performance policies effective in 2012. For a clean comparison, I limit the sample to hospitals who always have clinical executives or never have clinical executives. That is, these hospitals do not change their propensity to hire a clinical executive during the sample period. \footnote{An important assumption for this analysis is that this sample selection is not correlated with the program enactment. I investigate this assumption in Section \ref{sec:changes}, and find no evidence that hospitals are more or less likely to change their propensity to hire a clinical executive because of the pay-for-performance programs. }

@@

    In the unweighted two-way-fixed-effects specification, we have reason to believe that the parallel trends assumption may not hold due to observed differences in characteristics other than the leadership team, as discussed in Section \ref{sec:data}. Once incorporating the weights used in the synthetic differences in differences method, we ensure parallel pre-trends in the outcome variables. Thus, I assume that under the weighted clinical and non-clinical hospital composition, trends would have continued to be parallel absent the incentive change in 2012. This includes an implicit assumption that no other event occurred in 2012 correlated with both the outcomes and composition of leadership team. Finally, I assume that there is no anticipation of the policy change before 2012. The rules of the policies were announced in October of 2011, so 2012 was the first full year that hospitals could meaningfully respond to the policies.

    The estimated difference in readmission and mortality rates between hospitals with and without clinical executives are shown in Figure \ref{fig:clinicalsynthdid}. Estimate $\beta$ is represented in the difference in slope after 2012 when pay-for-performance incentives began. Panel \ref{fig:read_synth_clinical} shows the estimated difference in readmission rates, and reveals that non-clinical teams decrease readmissions by .3 (standard error .13) more than clinical teams. This is a relatively small magnitude given an average readmission rate of 21, but reveals a meaningful difference in behavior given that both hospital types significantly decrease readmissions. In panel \ref{fig:mort_synth_clinical}, I show the estimated difference in mortality. While the magnitude of the difference is similar to that for readmission rates, mortality is a noisy measure, and I cannot rule out null effects for mortality. In Appendix \ref{app:condition}, I show that these results are primarily driven by pneumonia patients.

     \begin{figure}[ht!]
     \caption{Effect of Clinical Training on Readmission and Mortality Rates}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{Readmission Rate}
         \includegraphics[width=\textwidth]{Objects/read_md_nomd_synth_graph.pdf}
         \label{fig:read_synth_clinical}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{Mortality Rate}
         \includegraphics[width=\textwidth]{Objects/mort_md_nomd_synth_graph.pdf}
         \label{fig:mort_synth_clinical}
     \end{subfigure}
        \label{fig:clinicalsynthdid}
    \end{figure}

    Additionally, I estimate the intensive margin of clinical leadership: whether the fraction of clinical executives on the team affects quality. I categorize hospitals with clinically trained executives into two bins: above the median and below median fraction of clinically trained executives. Let the median fraction be $\eta$, which in this case is .2. Then, I estimate the synthetic difference-in-differences version of the following specifications:

    \begin{equation}
    \label{eq:clinical_continuous_onefourth}
    y_{ht} = \beta \left(\mathbbm{1}\{\rho \leq \eta\}\text{ x post 2012}\right)_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht}, \text{ and}
    \end{equation}
    \begin{equation}
    \label{eq:clinical_continuous_onehalf}
    y_{ht} = \beta \left(\mathbbm{1}\{\rho > \eta\}\text{ x post 2012}\right)_{ht} + \gamma_{h} + \delta_t + \epsilon_{ht},
    \end{equation}

@@

    I present the results of this estimation with readmission rates as the outcome in Figure \ref{fig:cont_clinicalsynthdid}. In panel \ref{fig:belowmed_read_synth_clinical}, the estimation compares executive teams with no clinical training to those with up to 25\% of the team having clinical training. The average treatment effect is .04 and not statistically significant. Thus, there is no significant difference in how these types of teams respond to the incentive change. In panel \ref{fig:abovemed_read_synth_clinical}, I compare hospitals with no clinically trained executives over the course of the sample to hospitals with over a quarter of their team having clinical training. I find that hospitals without clinically trained executives decrease readmissions by .64 (standard error .33) more than hospitals with clinically trained executives. Thus, the readmission rate response is fully driven by having a higher proportion of clinical executives relative to no clinical executives.

    \begin{figure}[ht!]
     \caption{Effect of Clinical Training on Readmission Rates, Binned Treatment}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{$\rho \leq \eta$}
         \includegraphics[width=\textwidth]{Objects/cont_belowmedread_md_nomd_synth_graph.pdf}
         \label{fig:belowmed_read_synth_clinical}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{$\rho > \eta$}
         \includegraphics[width=\textwidth]{Objects/cont_abovemedread_md_nomd_synth_graph.pdf}
         \label{fig:abovemed_read_synth_clinical}
     \end{subfigure}
        \label{fig:cont_clinicalsynthdid}
    \end{figure}

    Any potential difference in mortality rate response also seems to be driven by executive teams with over a quarter of the executives having clinical training, as shown in Figure \ref{fig:cont_clinicalsynthdid_mort}. However, the standard errors of this estimation are high, potentially due to the noisiness in the mortality measure. The magnitude of the estimate in panel $\ref{fig:belowmed_mort_synth_clinical}$ is .07, so I can rule out differential responses by teams with less than a quarter of clinical executives. Taking the readmission and mortality rate results together, the effect of having clinical leadership on quality after an incentive change depends on having a sufficient proportion of the team with clinical training.

    \begin{figure}[ht!]
     \caption{Effect of Clinical Training on Mortality Rates, Binned Treatment}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{$\rho \leq \eta$}
         \includegraphics[width=\textwidth]{Objects/cont_belowmedmort_md_nomd_synth_graph.pdf}
         \label{fig:belowmed_mort_synth_clinical}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \caption{$\rho > \eta$}
         \includegraphics[width=\textwidth]{Objects/cont_abovemedmort_md_nomd_synth_graph.pdf}
         \label{fig:abovemed_mort_synth_clinical}
     \end{subfigure}
        \label{fig:cont_clinicalsynthdid_mort}
    \end{figure}
